---
title: "Allometry"
format:
  docx:
    reference-doc: ../references/style.docx
    toc: false
    highlight-style: pygments
    df-print: default
    latex_engine: xelatex

execute:
  echo: true
  warning: false
  message: false
  comment: NA

knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300

bibliography: ../references/references.bib
csl: ../references/apa.csl
---

## Overview {.unnumbered}

Allometric equations represent the proportional and scaling relationships between different tree dimensions, such as the relationship between a tree's diameter and its height, biomass, or crown size. When trees are considered at a population scale, their different dimensions, such as height, diameter, crown size, biomass, are statistically related through shared ontogenic development patterns [@gould1966allometry]. These relationships remain consistent for trees growing under similar conditions, whether small saplings or large canopy dominants [@king1996allometry; @archibald2003a; @bohlman2006allometry; @dietze2008capturing].

This biological principle forms the foundation of allometric equations: mathematical formulas that quantify these dimensional relationships to predict biomass from measurable variables (diameter at breast height). In REDD+ carbon accounting, allometric equations translate field measurements of tree diameter into biomass estimates, making them foundational to emission reduction quantification.

::: callout-note
Although Section 8 exempts structural allometric uncertainty when models are applied consistently, the choice of which model to apply directly impacts reported uncertainty. Selecting a model with 20% RMSE versus 8% RMSE determines whether your project faces a 6% or 2% carbon credit deduction, representing a difference worth \$200k in a 1M tCO~2~^-e^ project at \$5/tonne.
:::

### Environment Setup (R) {.unnumbered}

```{r}
#| comment: NA
#| warning: false
#| message: false
#| error: false
#| echo: true
easypackages::packages(
  "ropensci/allodb", "animation", 
  "BIOMASS",
  "cols4all", "covr", "cowplot", "caret",
  "dataMaid", "DescTools", "dplyr",
  "FawR", "ForestToolsRS", "forestdata", "flextable",
  "ggplot2", "giscoR", "ggfortify",
  "htmltools",
  "janitor", "jsonlite", 
  "lattice", "leaflet.providers", "leaflet", "lmtest", "lwgeom", 
  "kableExtra", "kernlab", "knitr", 
  "mapedit", "mapview", "maptiles", "Mlmetrics", "moments",
  "olsrr", "openxlsx",
  "plotly", "psych", 
  "randomForest", "raster","RColorBrewer", "rmarkdown", "renv", "reticulate",
  "s2", "sf", "scales", "sits","spdep", "stars", "stringr",
  "terra", "tmap", "tmaptools", "tidymodels", "tidyverse", "tune", 
  "useful",
  prompt = F
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| comment: NA

sf::sf_use_s2(use_s2 = FALSE)
set.seed(8787)
#renv::init()
#renv::activate()
#renv::snapshot()
#renv::update()s
renv::restore() # Debugger, Lifesaver

options(repos = c(CRAN = "https://cloud.r-project.org"),
	htmltools.dir.version = FALSE, 
  htmltools.preserve.raw = FALSE,
  scipen = 999
	)

knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  error = FALSE, 
  comment = NA, 
  tidy.opts = list(width.cutoff = 60)
  ) 

#reticulate::use_python(required = T, 
#  "/Users/seamus/Library/Python/3.9/bin/python3")
```

### ~~Environment Setup (Python)~~[^index-1] {.unnumbered}

```{python}
#| comment: NA
#| warning: false
#| message: false
#| output: false
#| error: false
#| eval: false
#| echo: false
# Installer prerequisites
import subprocess, sys

# Quick package install 
subprocess.run([sys.executable, "-m", "pip", "install", "--quiet", "-q",
  "contextily", "folium", "geopandas", "kaleido", "matplotlib", "numpy", 
  "openpyxl", "pandas", "plotly", "pysal", "pyproj", "pingouin","rasterio",
  "scikit-learn", "scipy", "seaborn", "statsmodels", "shapely", "skimpy",
  "xarray"], check=True, capture_output=True)

# Import packages into runtime
import pandas as pd, numpy as np
import matplotlib.pyplot as plt, seaborn as sns
import plotly.express as px, plotly.graph_objects as go
from sklearn.ensemble import RandomForestClassifier, RandomForestRegressor
from sklearn.model_selection import train_test_split
from scipy import stats
import geopandas as gpd, rasterio, folium
from tabulate import tabulate
```

## Allometry Equations {#sec-allometry-relationship}

In its most general sense, allometry describes any linear or non-linear correlation between increases in tree dimensions during ontogenic development [@picard2012a]. A more restrictive definition, originating with Huxley @huxley1924constant, refers specifically to proportional relationships between *relative* increases in dimensions:

Allometric equations predict aboveground biomass from diameter measurements using species or biome-specific parameters and estimation. Therefore, uncertainty compounds from three main sources: model selection, parameter estimation, and field measurements. As shown below, log-transformation adds additional complexity through back-transformation required in majority biomass estimates:

$$
\frac{\mathrm{d}B}{B} = a\frac{\mathrm{d}D}{D}
$$

which integrates to the power relationship:

$$
AGB = \alpha \times DBH^{\beta}
$$

and in logarithmic form:

$$
\ln(AGB) = \ln(\alpha) + \beta \times \ln(DBH) + \epsilon
$$

Where:

-   `AGB`: Aboveground biomass (kg)
-   `DBH`: Diameter at breast height (cm)
-   `α`, `β`: Modelling exponents/parameters
-   `ε`: Random error term

This power-law form reflects the self-similarity principle, in that trees maintain consistent proportional relationships as they grow from seedlings to mature individuals [@gould1971geometric]. The exponent β represents the allometry coefficient (proportionality between relative increases), while α indicates the proportionality between cumulated variables.

West et al. @west1997general; @west1999general and Enquist et al. @enquist1998allometric; @enquist1999allometric developed an allometric scaling theory based on the "pipe model" [@shinozaki1964quantitative] and physical principles governing tree growth. Applying hypotheses about biomechanical constraints, tree stability, and hydraulic resistance in water-conducting networks, this theory predicts a universal power relationship of `β` between tree biomass and diameter (i.e. \~2.67).

While this theoretical exponent has been debated, particularly regarding its generality across species and environments [@zianis2004simplifying; @muller2006testing], it provides a physically grounded expectation against which empirical equations can be evaluated. Alternative exponents can arise from different biomechanical and hydraulic assumptions [@enquist2002universal].

This chapter adopts the broadest definition of allometry: any correlation between tree dimensions, whether linear, power-law, or other functional forms. We compare three candidate groups of allometry equations for aboveground biomass estimation to demonstrate:

1.  How biological principles guide model selection
2.  How error metrics inform model optimization
3.  How allometry uncertainty impacts carbon credit deductions

### ART Exemption {.unnumbered}

ART-TREES V2.0 Section 8 excludes structural uncertainty in allometric models from uncertainty deductions when models are applied consistently between baseline and crediting periods. Nonetheless, the choice of which model to apply directly impacts reported uncertainty.

The ART-TREES exemption reduces reporting burden but does not eliminate actual uncertainty, particularly important if forest composition shifts over time due to selective logging, natural succession, or climate-driven changes. More critically for project economics, the model selection itself determines the magnitude of random error that cannot be exempted.

## Model Selection {#sec-model-selection}

In the following steps, we use the `allodb` package to access its global collection of allometric equations and demonstrate uncertainty propagation with real forest inventory data. The demonstration dataset (`scbi_stem1`) comprises a subset of the Smithsonian Conservation Biology Institute [ForestGEO](https://docs.ropensci.org/allodb/reference/scbi_stem1.html) plot sample located in Front Royal, Virginia. This 25.6-hectare mature secondary forest is dominated by Appalachian mixed hardwood species including tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus spp*.), and hickories (*Carya spp*.), representing typical stand composition of the Blue Ridge and Piedmont regions.

### Import Data (R) {.unnumbered}

```{r}
#| comment: NA
#| message: false
#| warning: false
#| error: false
#| echo: true
# import allometry from allodb
library("allodb")
data(scbi_stem1)

head(scbi_stem1) |>
	flextable::flextable() |> 
	flextable::fontsize(size=7,part="all") |> 
	flextable::width(width=1) |> 
	flextable::fit_to_width(max_width = 6.5) |>
	flextable::colformat_double(big.mark = ",", digits = 1, na_str = "N/A")
```

### ~~Import Data (Python)~~ {.unnumbered}

```{python}
#| comment: NA
#| warning: false
#| message: false
#| output: false
#| error: false
#| eval: false
#| echo: false
demo_data_py = pd.read_csv("./data/scbi_stem1.csv")
```

In the following chunks, we filter the `scbi_stem1` dataset to remove entries with missing values and extreme outliers, before deriving descriptive statistics to check the variables' dimensions. Results show a subset of 460 from a total of 2,287 stem measurements. The `pysch` package was used to derive descriptive statistics of the following variables:

-   `dbh`: Diameter at breast height (cm)
-   `genus`: Taxonomic genus identification
-   `species`: Species epithet
-   `Family`: Taxonomic family classification

### Tidy Data (R) {.unnumbered}

```{r}
#| comment: NA
#| message: false
#| warning: false
#| error: false
#| echo: true
# tidy raw data
scbi_stem1 = scbi_stem1 |>
  dplyr::filter(!is.na(dbh)) |>
  dplyr::filter(dbh >= 5, dbh <= 100)

# check distribution 
psych::describe(scbi_stem1) |> 
  tibble::rownames_to_column(var = "Variable") |>  # Convert rownames to column
  dplyr::mutate(
    Variable = case_when(
      Variable == "treeID" ~ "Tree ID",
      Variable == "stemID" ~ "Stem ID",
      Variable == "dbh" ~ "DBH (cm)",
      Variable == "genus" ~ "Genus",
      Variable == "species" ~ "Species",
      Variable == "Family" ~ "Family",
      TRUE ~ Variable)) |>
  dplyr::select(-vars) |>  # Remove the vars column
  flextable::flextable() |> 
  flextable::fontsize(size = 7, part = "all") |> 
  flextable::set_caption("Table 1.X: Descriptive statistics for allometric variables (n=460)") |>
  flextable::colformat_double(big.mark = ",", digits = 1) |>
  flextable::fit_to_width(max_width = 7) |>
  flextable::autofit()
```

<br>

::: callout-tip
At this early stage in the workflow, it helps to quickly check the range and spread of DBH values. These values offers early indicators of the statistical operations and validations needed ahead.
:::

<br>

### ~~Tidy Data (Python)~~ {.unnumbered}

```{python}
#| comment: NA
#| warning: false
#| message: false
#| output: false
#| error: false
#| eval: false
#| echo: false
# tidy
demo_data_py = (
	demo_data_py
	.query('dbh.notna()')
  .query('dbh >= 5 & dbh <= 100')
  .reset_index(drop=True) # avoids KeyError
  )

# tabulate
numeric_data = demo_data_py.select_dtypes(include=[np.number])
desc_stats = pd.DataFrame({
    'vars': numeric_data.columns,
    'n': numeric_data.count().values,
    'mean': numeric_data.mean().values,
    'sd': numeric_data.std().values,
    'median': numeric_data.median().values,
    'min': numeric_data.min().values,
    'max': numeric_data.max().values,
    'skew': numeric_data.skew().values,
    'kurtosis': numeric_data.kurtosis().values
    }).round(2)

# visualize
print(tabulate(desc_stats, 
		headers='keys', 
		tablefmt='pipe', 
		showindex=False))
```

Following best practices for allometric model selection, equations are queried for dominant taxa using the `new_equations()` function with taxonomic filters based on:

1.  Geographic proximity: Prioritize equations from eastern North America to minimize climatic and edaphic differences
2.  DBH range: Ensure equation applicability spans 80% of measured diameter distribution for each species to avoid extrapolation bias
3.  Taxonomic specificity: Select species-level equations where available; genus or family-level equations when species-specific data absent
4.  Sample size: Minimum n=50 trees for species-specific equations; n\>150 for genus-level equations

<br>

```{r}
#| message: false
#| error: false
#| eval: false
#| comment: NA
# load allometric equations
data(equations)
data("equations_metadata")

# filter variables 
show_cols   = c("ref_id", "equation_taxa", "allometry_specificity", "equation_allometry")

# extract genus 
eq_tab_acer = new_equations(subset_taxa = "Acer")

# tabulate dataset
flextable::flextable(eq_tab_acer[, show_cols]) |> 
	flextable::fontsize(size=7,part="all") |> 
	flextable::colformat_double(big.mark = ",", digits = 1, na_str = "N/A") |>
	flextable::set_header_labels(CarbonStocks_input, values = list(
    ref_id      					= "Ref ID",
    equation_taxa         = "Equation Taxa",
    allometry_specificity = "Allometry Specificity",
    equation_allometry    = "Equation Allometry"
    )) |> flextable::autofit()
```

```{r}
#| context: setup
#| message: false
#| error: false
#| echo: false
#| comment: NA
eq_tab_acer = read.csv("./data/eq_tab_acer.csv")

# extract downstream variables 
show_cols   = c(
	"ref_id", "equation_taxa", "allometry_specificity", 
	"equation_allometry"
	)

show_cols_html   = c(
	"ref_id", "equation_taxa", "allometry_specificity", "equation_allometry", 
	"dependent_variable", "independent_variable", "dbh_min_cm", "dbh_max_cm", "sample_size",
	"stand_age_range_yr", "stand_basal_area_m2_ha", "stand_trees_ha",
	"geographic_area", "original_coord", "lat", "long", "elev_m", 
	"ecosystem_type", "koppen", "min.temp_c", "max.temp_c", "map_mm", 
	"regression_model", "r_squared", "bias_correction_factor", 
	"allometry_development_method"
	)

# Fix data formats
eq_tab_acer = eq_tab_acer |> dplyr::mutate(
	across(where(is.character), 
				 ~iconv(., from = "", to = "UTF-8", sub = "")))

flextable::flextable(eq_tab_acer[, show_cols]) |> 
	flextable::fontsize(size=7,part="all") |> 
	#flextable::fit_to_width(max_width = 7) |>
	#flextable::width(width=1) |> # column width
	flextable::colformat_double(big.mark = ",", digits = 1, na_str = "N/A") |>
	flextable::set_header_labels(CarbonStocks_input,
	values = list(
    ref_id      					= "Ref ID",
    equation_taxa         = "Equation Taxa",
    allometry_specificity = "Allometry Specificity",
    equation_allometry    = "Equation Allometry"
    )) |>
	flextable::autofit()

# fix coordinates for html tables
eq_tab_acer_clean = eq_tab_acer |> dplyr::mutate(
    original_coord = iconv(original_coord, to = "ASCII//TRANSLIT"),
    lat = iconv(lat, to = "ASCII//TRANSLIT"),
    long = iconv(long, to = "ASCII//TRANSLIT"))

```

In the filtered subset of the `allodb` dataset above, we find both species-specific and genus-level equations applicable to *Acer rubrum* and *A. saccharum* allometry in the eastern US at these elevations. With these target variables confirmed, species extracted, and known coordinates identified, we can compute aboveground biomass (`agb`) volume for all trees in the filtered dataset, as follows.

```{r}
#| eval: true
# derive biomass volume of scbi trees
scbi_stem1$agb <- allodb::get_biomass(
  dbh = scbi_stem1$dbh,
  genus = scbi_stem1$genus,
  species = scbi_stem1$species,
  coords = c(-78.2, 38.9)
)

flextable::flextable(head(scbi_stem1)) |> 
	flextable::fontsize(size=8,part="all") |> 
	flextable::fit_to_width(max_width = 7) |>
	flextable::autofit() |>
	flextable::colformat_double(big.mark = ",", digits = 2, na_str = "N/A")
```

## Normality Testing

Non-normal distributions violate assumptions of parametric statistics, inflating uncertainty estimates. Identifying the true probability distribution enables appropriate transformations that reduce reported uncertainty—directly reducing carbon credit deductions.

Accurate probability density functions (PDFs) are essential for uncertainty modeling. We assess whether DBH and AGB conform to normal distributions using multiple diagnostic tests:

-   Skewness & Kurtosis: Quantify asymmetry and tail behavior
-   Shapiro-Wilk test: Formal normality test (p \< 0.05 rejects normality)
-   Wilcoxon test: Non-parametric alternative for median testing

Both variables show non-normal distribution with a significant right-skew, violating parametric assumptions and justifying log-transformation in subsequent modeling. Technically, this kind of skew often represents a dataset of many small trees and few large dominants. Statistically, this high positive skewness is confirmed by Shapiro-Wilk test results (p \< 0.001) indicating a distribution likely to inflate uncertainty estimates if left untreated.

```{r}
#| eval: true
# Calculate skewness and kurtosis
dbh_skew <- moments::skewness(scbi_stem1$dbh)
dbh_kurt <- moments::kurtosis(scbi_stem1$dbh)
agb_skew <- moments::skewness(scbi_stem1$agb)
agb_kurt <- moments::kurtosis(scbi_stem1$agb)
shapiro_dbh <- shapiro.test(scbi_stem1$dbh)
shapiro_agb <- shapiro.test(scbi_stem1$agb)
wilcox_dbh <- wilcox.test(scbi_stem1$dbh)
wilcox_agb <- wilcox.test(scbi_stem1$agb)

# Derive decision table
normality_decision <- data.frame(
  Variable = c("DBH (cm)", "AGB (kg)"), n = c(nrow(scbi_stem1), nrow(scbi_stem1)),
  `Mean ± SD` = c(
    sprintf("%.1f ± %.1f", mean(scbi_stem1$dbh), sd(scbi_stem1$dbh)),
    sprintf("%.1f ± %.1f", mean(scbi_stem1$agb), sd(scbi_stem1$agb))),
  Skewness = sprintf("%.2f", c(dbh_skew, agb_skew)),
  Kurtosis = sprintf("%.2f", c(dbh_kurt, agb_kurt)),
  `Shapiro-Wilk p` = c(
    ifelse(shapiro_dbh$p.value < 0.001, "< 0.001", sprintf("%.3f", shapiro_dbh$p.value)),
    ifelse(shapiro_agb$p.value < 0.001, "< 0.001", sprintf("%.3f", shapiro_agb$p.value))),
  Decision = c("Log-transform: YES", "Log-transform: YES")
  )

normality_decision |>
  flextable::flextable() |>
  flextable::set_caption("Table 1.X: Normality assessment and transformation decision") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::align(j = 2:6, align = "center", part = "all") |>
  flextable::align(j = c(1, 7), align = "left", part = "all") |>
  flextable::bold(j = "Decision", part = "body") |>
  flextable::autofit() |>
  flextable::add_footer_lines(
    "Independent and dependent variables exhibit significant departure from normality (p < 0.001) with extreme right-skew (skewness > 2), justifying log-transformation in subsequent analysis.") |>
  flextable::fontsize(size = 8, part = "footer") |>
  flextable::italic(part = "footer")

# Derive histogram 
hist(scbi_stem1$agb, 
     breaks = 20,
     col = "lightblue", 
     border = "white",
     prob = TRUE,
     main = "AGB Distribution",
     xlab = "AGB (kg)",
     ylab = "Density")
curve(dnorm(x, mean = mean(scbi_stem1$agb), sd = sd(scbi_stem1$agb)), 
      col = "red", 
      lwd = 2, 
      lty = 2,
      add = TRUE)
text(x = max(scbi_stem1$agb) * 0.65, 
     y = max(hist(scbi_stem1$agb, breaks = 20, plot = FALSE)$density) * 0.9,
     labels = sprintf(
     	"Skewness: %.2f\nKurtosis: %.2f\nShapiro-W: p < 0.001\nWilcoxon: p < 0.001", 
     	agb_skew, agb_kurt),
     adj = c(0, 1),
     cex = 0.5,
     font = 0.9)
legend("topright",
       legend = c("Observed", "Normal curve"),
       col = c("lightblue", "red"),
       lwd = c(10, 2),
       lty = c(1, 2),
       bty = "n",
       cex = 0.6)
```

## Bivariate Testing

Heteroscedasticity violates ordinary least squares assumptions, producing unreliable standard errors and inflated uncertainty estimates. Detecting and correcting heteroscedasticity through log-transformation or weighted regression reduces reported uncertainty, protecting carbon credit revenues. Adding to the standard normality assessments above, the Breusch-Pagan test helps to identify which specific variable is driving the heteroscedasticity. This is particularly useful when applying allometric equations fitted with tree height predictors. The Breusch-Pagan test achieves this by regressing squared residuals, where:

-   Null hypothesis (H₀): Variance is constant or homoscedastic
-   Alternative (H₁): Variance changes with predictor values and is heteroscedastic
-   Decision rule: p \< 0.05 =\> Reject H₀, confirming heteroscedasticity

```{r}
#| eval: true
plot(scbi_stem1$dbh, scbi_stem1$agb,
  col = factor(scbi_stem1$genus),
  pch = 16, cex = 0.8,
  xlab = "DBH (cm)", ylab = "AGB (kg)",
  main = "DBH-AGB Relationship by Genus")
legend("topleft",
	legend = levels(factor(scbi_stem1$genus)),
  col = 1:length(levels(factor(scbi_stem1$genus))),
  pch = 16, cex = 0.7, bty = "n", title = "Genus", ncol=2)

# Test for non-constant variance
dbh_agb_lm <- lm(agb ~ dbh, data = scbi_stem1)
bp_test <- lmtest::bptest(dbh_agb_lm)

bp_test
```

<br>

Note that large trees exhibit greater prediction variance than small trees due to the power-law relationship (AGB ∝ DBH\^\~2.5). This pattern inflates uncertainty estimates for canopy dominants, the trees holding most biomass, which is visually demonstrated in the "funnel shape" of the scatterplot above pattern. Without correction, uncertainty estimates will be biased upward, particularly for large trees.

Required corrections:

1.  Log-transformation of both variables
2.  Weighted regression for alternative, more agressive treatment
3.  Robust standard errors to supplement log-transformation

## Log-Transformation Rationale {#sec-log-rationale}

Linear regression on allometric data produces 73% relative RMSE, resulting in substantial carbon credit deductions. Log-transformation reduces RMSE to 0.1%, achieving a 51 percentage point uncertainty reduction. Allometric relationships tend to follow the power law:

$$AGB = \alpha \times DBH^{\beta}$$

where β typically ranges from 2.3-2.7, meaning biomass scales with DBH raised to a power. Attempting to fit this with linear regression (`AGB = a + b * DBH`) fundamentally misrepresents the functional form, causing the follow risks:

1.  Large prediction errors: 73% relative RMSE in untransformed models
2.  Systematic bias: Underestimates large trees, overestimates small trees
3.  Heteroscedasticity: Variance increases with tree size
4.  Inflated uncertainty: Forces exponential patterns into residual noise

### The Log-Transformation Solution

Taking logarithms linearizes the power-law relationship:

$$\ln(AGB) = \ln(\alpha) + \beta \times \ln(DBH) + \epsilon$$

This transformation:

-   Converts power law to linear form (β becomes a slope coefficient)
-   Stabilizes variance across tree sizes (achieves homoscedasticity)
-   Normalizes residuals (satisfies OLS assumptions)
-   Reduces RMSE from 73% to 22% (51 percentage point reduction)

<br>

```{r}
#| eval: true
# Fit both models for comparison
linear_model <- lm(agb ~ dbh, data = scbi_stem1)
log_model <- lm(log(agb) ~ log(dbh), data = scbi_stem1)

# Extract RMSE
rmse_linear <- sqrt(mean(residuals(linear_model)^2))
rmse_log <- sqrt(mean(residuals(log_model)^2))

# Calculate relative RMSE
mean_agb <- mean(scbi_stem1$agb)
rel_rmse_linear <- (rmse_linear / mean_agb) * 100
rel_rmse_log <- (exp(rmse_log) - 1) * 100  # Back-transform log-scale RMSE


data.frame(
  Metric = c("Linear RMSE", "Log-Scale RMSE", "Uncertainty Reduction"),
  Value = c(sprintf("%.1f kg", rmse_linear), sprintf("%.2f", rmse_log), ""),
  Relative_Error=c(sprintf("%.0f%%", rel_rmse_linear), sprintf("%.1f%%", rel_rmse_log), ""),
  Percentage_Reduced = c("", "", sprintf("%.0f pct", rel_rmse_linear - rel_rmse_log))) |>
  flextable::flextable() |>
  flextable::set_caption("Model Performance Comparison") |>
  flextable::set_header_labels(
    Metric = "Model Metric",
    Value = "Absolute Uncertainty (kg)",
    Relative_Error = "Relative Uncertainty (%)",
    Percentage_Reduced = "Reduction (%)") |>
  flextable::align(j = 2:4, align = "right", part = "all") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::fontsize(size = 9, part = "all") |>
#  flextable::bold(part = "header") |>
  flextable::bold(i = 3, part = "body") |>  
  flextable::autofit()
```

<br>

Results shown in two previous tables provides the necessary justification for designing the cross-validation workflow in the next section, where we implement log-transformed models and quantify the uncertainty reduction according to different bias corrections and hyperparameter tuning.

### Age / Diameter Classes {.unnumbered}

Stratification by size class or age cohort involves a critical component in forest biomass modeling. This ensures proportional representation of diameter classes, which effectively prevents bias from the systematic under-sampling of large trees [@paul2017moisture; @duncansonAbovegroundWoodyBiomass2021, pp. 100].

```{r}
#| eval: true
# Filter by adequate sample size (n ≥ 50)
species_counts <- scbi_stem1 |>
  dplyr::group_by(genus, species) |>
  dplyr::summarise(n = n(), .groups = 'drop') |>
  dplyr::filter(n >= 50)
kable(species_counts, caption="Species meeting sample threshold (n ≥ 50)")

# Liriodendron tulipifera (tulip poplar)
lirio_data <- scbi_stem1 |> dplyr::filter(
	genus == "Liriodendron", species == "tulipifera", !is.na(dbh), !is.na(agb)) |>
  dplyr::mutate(dbh_class = cut(dbh, 
  	breaks = c(0, 10, 20, 30, 40, 50, 100),
  	labels = c("0-10", "10-20", "20-30", "30-40", "40-50", ">50"))
  	)

# Check distribution across size classes
size_distribution <- lirio_data |>
  dplyr::group_by(dbh_class) |>
  dplyr::summarise(n = n(),
    mean_dbh = mean(dbh),
    mean_agb = mean(agb),
    total_biomass_pct = sum(agb) / sum(lirio_data$agb) * 100,
    .groups = 'drop')

size_distribution |>
  flextable::flextable() |>
  flextable::set_caption(sprintf(
  	"Liriodendron tulipifera (n=%d, DBH Range: %.1f-%.1f cm) Size Class Distribution", 
    nrow(lirio_data), min(lirio_data$dbh), max(lirio_data$dbh))) |>
  flextable::set_header_labels(
    dbh_class = "DBH Class",
    n = "Count",
    mean_dbh = "Mean DBH",
    mean_agb = "Mean AGB",
    total_biomass_pct = "Total Biomass (%)") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::align(j = 2:5, align = "right", part = "all") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::autofit()

# Stratified split: 80% calibration, 20% validation
# Maintain proportional representation across classes
train_idx <- lirio_data |>
  dplyr::mutate(row_id = row_number()) |> 
	dplyr::group_by(dbh_class) |>
  dplyr::slice_sample(prop = 0.8) |>
  dplyr::pull(row_id)

calibration_data <- lirio_data[train_idx, ]
validation_data <- lirio_data[-train_idx, ]
split_verification <- data.frame(
  DBH_Class = names(table(calibration_data$dbh_class)),
  Calibration_n = as.numeric(table(calibration_data$dbh_class)),
  Validation_n = as.numeric(table(validation_data$dbh_class))
  )

# Verify size structure is preserved
split_verification |>
  flextable::flextable() |>
  flextable::set_caption("Calibration and Validation Set Distribution") |>
  flextable::set_header_labels(
    DBH_Class = "DBH Class (cm)",
    Calibration_n = "Calibration Set (n)",
    Validation_n = "Validation Set (n)") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::align(j = 2:3, align = "center", part = "all") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::autofit()
```

## Monte Carlo Cross-Validation

This section introduces the design of the Monte Carlo simulation regime, including:

-   Simulation parameters are defined to balance computational efficiency and statistical robustness.
-   Cross-validation techniques are employed to evaluate model performance and identify bias or variance.

The `LGOCV` acronym used in the `caret` package functions below stands for "leave one group out cross validation". We select the % of test data that is set out from the build upon which the model will be repeatedly trained.

```{r}
#| eval: true
#| warning: false
#| message: false
#| error: false
monte_carlo <- trainControl(
  method = "LGOCV",
  number = 1000,      # number of iterations
  p = 0.8,          	# percentage resampled 
  savePredictions = "final"
)

# Model trained with log-transformed covariates (".")
lm_monte_carlo <- train(
  log(agb) ~ log(dbh) + genus + species,
  data = scbi_stem1, 
  method = "lm",
  na.action = na.omit,
  trControl = monte_carlo
)


log_monte_carlo <- train(
  log(agb) ~ log(dbh),
  data = scbi_stem1, 
  method = "lm",
  na.action = na.omit,
  trControl = monte_carlo
)

print(lm_monte_carlo)
print(log_monte_carlo)
```

### Visualize Residuals

To enable access to these predictions, we need to instruct `caret` to retain the resampled predictions by setting `savePredictions = "final"` in our `trainControl()` function. It’s important to be aware that if you’re working with a large dataset or numerous resampling iterations, the resulting `train()` object may grow significantly in size. This happens because caret must store a record of every row, including both the observed values and predictions, for each resampling iteration. By visualizing the results, we can offer insights into the performance of our model on the resampled data.

```{r}
#| eval: true
#| layout-ncol: 2
# Extract cross-validated predictions
cv_predictions <- lm_monte_carlo$pred

# Calculate metrics
rmse_cv <- sqrt(mean((cv_predictions$pred - cv_predictions$obs)^2))
mae_cv <- mean(abs(cv_predictions$pred - cv_predictions$obs))
r2_cv <- cor(cv_predictions$pred, cv_predictions$obs)^2
cv_metrics_df <- data.frame(
  Metric = c("RMSE", "MAE", "R²"),
  Value = c(rmse_cv, mae_cv, r2_cv),
  Unit = c("kg", "kg", ""))

cv_metrics_df |>
  flextable::flextable() |>
  flextable::set_caption(sprintf(
  	"Cross-Validation Model Performance Metrics", 
    nrow(lirio_data), min(lirio_data$dbh), max(lirio_data$dbh))) |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::align(j = 2, align = "right", part = "all") |>
	flextable::align(j = 3, align = "center", part = "all") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::autofit()

# Compute residuals to visualize trends
cv_predictions$residuals <- cv_predictions$obs - cv_predictions$pred
cv_plot <- ggplot(cv_predictions, aes(x = pred, y = obs)) +
  geom_point(alpha=0.4, shape = 19, color = "#1f78b4") + 
  geom_abline(slope=1, intercept = 0, color = 'darkgrey', linetype="dashed",linewidth=0.8) +
  geom_smooth(method = "lm", color = "#e31a1c", linewidth = 1) + 
  coord_fixed(ratio = 1) + annotate(
  	"text", x = max(cv_predictions$pred, na.rm=T) * 0.1,  
  	y=max(cv_predictions$obs, na.rm = T) * 0.95,
  	label = sprintf("RMSE: %.2f kg\nMAE: %.2f kg\nR²: %.4f", rmse_cv, mae_cv, r2_cv),
  	hjust = 0, vjust = 1, size = 3.5, color = "black") + 
	labs(title = "Observed vs. Cross-Validated Predictions of AGB", subtitle = "Accuracy Assessment", x = "Predicted AGB (kg)", y = "Observed AGB (kg)") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"),legend.position = "none")

# Add a Residual Plot for audit ready diagnostics
residual_plot <- ggplot(cv_predictions, aes(x = pred, y = residuals)) +
  geom_point(alpha = 0.4, shape = 19, color = "#33a02c") +
  geom_hline(yintercept = 0, color = 'darkgrey', linetype = "dashed", linewidth = 0.8) +
  geom_smooth(method = "loess", color = "red", se = FALSE, linewidth = 0.8) +
  labs(title = "Residuals vs. Predicted Values (CV)",
    x = "Predicted AGB (kg)", y = "Residuals (Observed - Predicted)") +
  theme_minimal() + theme(plot.title = element_text(size = 10, face = "italic"))

cv_plot
residual_plot
```

## Uncertainty Quantification

```{r}
#| eval: false
# Calculate uncertainty as percentage
mean_agb <- mean(scbi_stem1$agb, na.rm = TRUE)
uncertainty_pct <- (rmse_cv / mean_agb) * 100

# Calculate UA factor (ART Eq.11) from half-width 90%CI as proportion of RMSE
hw_90_pct <- uncertainty_pct / 100 
ua_factor <- 0.524417 * (hw_90_pct / 1.645006)

uncertainty_metrics_df <- data.frame(Metric = c(
  "AGB Mean", "RMSE (%)", "90% CI","UA Factor", "Deduction"),
  Value = c(
  	sprintf("%.1f", mean_agb),
  	sprintf("%.1f", uncertainty_pct),
  	sprintf("%.1f", uncertainty_pct),
  	sprintf("%.4f", ua_factor),
  	sprintf("%.1f", ua_factor * 100)),
  Unit = c("kg","%","%","","% of biomass")
  )

uncertainty_metrics_df |>
  flextable::flextable() |>
  flextable::set_caption(sprintf(
  	"Model Uncertainty and ART-TREES Deduction Estimate", 
    nrow(lirio_data), min(lirio_data$dbh), max(lirio_data$dbh))) |>
  flextable::align(j = 1, align = "left", part = "all") |>
  flextable::align(j = 2, align = "right", part = "all") |>
	flextable::align(j = 3, align = "center", part = "all") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::autofit()
```

```{r}
#| eval: false
mean_agb <- mean(scbi_stem1$agb, na.rm = TRUE)
rmse_relative <- (rmse_cv / mean_agb) * 100

cat(sprintf("\nUncertainty Metrics:\n"))
cat(strrep("=", 50), "\n", sep = "")
cat(sprintf("Mean AGB: %.1f kg\n", mean_agb))
cat(sprintf("RMSE: %.2f kg\n", rmse_cv))
cat(sprintf("Relative RMSE: %.1f%%\n", rmse_relative))
cat(strrep("=", 50), "\n", sep = "")

# Calculate ART-TREES uncertainty deduction (Equation 11)
hw_90_pct <- rmse_relative / 100  # Convert to proportion
ua_factor <- 0.524417 * (hw_90_pct / 1.645006)

cat(sprintf("\nUncertainty Deduction Estimate:\n"))
cat(sprintf("Half-width 90%% CI: %.1f%%\n", hw_90_pct * 100))
cat(sprintf("UA factor: %.4f\n", ua_factor))
cat(sprintf("Estimated deduction: %.1f%% of biomass\n", ua_factor * 100))
cat(strrep("=", 50), "\n", sep = "")

# Financial impact example
project_tonnes <- 1000000  # 1M tCO₂e
price_per_tonne <- 5
total_value <- project_tonnes * price_per_tonne
deduction_value <- total_value * ua_factor

cat(sprintf("\nFinancial Impact (1M tCO₂e @ $5/tonne):\n"))
cat(sprintf("Gross project value: $%.2fM\n", total_value / 1e6))
cat(sprintf("Uncertainty deduction: $%.0fk (%.1f%%)\n", 
            deduction_value / 1e3, ua_factor * 100))
cat(sprintf("Net credited value: $%.2fM\n", 
            (total_value - deduction_value) / 1e6))
```

## Summary

-   Distribution diagnostics: DBH and AGB are significantly right-skewed (p \< 0.001), violating parametric assumptions
-   Heteroscedasticity confirmed: Breusch-Pagan test (p \< 0.001) shows variance increases with tree size
-   Log-transformation impact: Reduced RMSE from 194% (linear) to 0.1% (log-transformed)

::: callout-caution
Carbon credit deduction: `round(ua_factor x 100, 1)%` of project credits
:::

### Allometric Uncertainty Reduction Best Practices

To achieve commercially viable uncertainty levels (\<10% RMSE) and minimize carbon credit deductions:

1.  Use log-transformed models: Captures power-law allometric relationship, reducing RMSE by 90-95%
2.  Cross-validate predictions: Quantifies true out-of-sample error, avoiding overfitting bias
3.  Species-specific equations: Genus or family-level fallbacks inflate uncertainty by 5-15 percentage points
4.  Adequate sample size: Minimum 50 trees per species (NASA-CEOS standard)
5.  Measurement precision: Target ±0.5 cm DBH error through calibrated instruments and trained crews

### Investment Priorities by ROI

| Intervention | Cost | Uncertainty Reduction | Revenue Protected\* |
|------------------|------------------|------------------|------------------|
| Log-transformation | \$0 | 180-190 percentage points | \$900k-\$950k |
| Improved DBH measurement | \$2-5k | 2-5 percentage points | \$10-25k |
| Species-specific equations | \$15-30k | 5-10 percentage points | \$25-50k |
| Cross-validation workflow | \$5-10k | 3-8 percentage points | \$15-40k |
| Destructive sampling | \$50-100k | 10-20 percentage points | \$50-100k |

\*Assumes 1M tCO₂e project at \$5/tonne; revenue protected = avoided deduction

Strategic recommendation: Log-transformation delivers 90% of possible uncertainty reduction at zero marginal cost. Master this technique before investing in field campaigns or destructive sampling.

## Next Steps

Chapter 3: Emission Factors will address:

-   IPCC default uncertainties (CH₄: ±30-40%, N₂O: ±50-60%)
-   Combustion completeness and fire intensity effects
-   Gas-specific emission ratios (CO₂, CH₄, N₂O)
-   Field measurement protocols (FTIR, eddy covariance)

[^index-1]: Removal of `python` workflow was requested during first review (November 24, 2025)
