---
title: "notes"
editor: visual
---

```{r setup}
#| warning: false
#| message: false
#| include: false
#| error: false
#| echo: false

library(remotes)
install_github("ytarazona/ForesToolboxRS")
suppressMessages(library(ForesToolboxRS))

#install.packages("pak")
easypackages::packages(
  'caret',
  'raster', 
  'sf', 
  'snow',
  'tmap',
  prompt = F
  )

knitr::opts_chunk$set(
  echo        = T, 
  message     = F, 
  warning     = F, 
  error       = T, 
  comment     = NA
  ) 

base::options(
  htmltools.dir.version  = F, 
  htmltools.preserve.raw = F)

# switch on/off for spherical geometries
sf::sf_use_s2(use_s2 = FALSE)
```

## Import AOI

```{r}
# assign master crs
crs_master    = sf::st_crs("epsg:4326")
# derive aoi windows
country   = geodata::gadm(country="GUY", level=0, path=tempdir()) |>
  sf::st_as_sf() |> sf::st_cast() |> sf::st_transform(crs_master)

states    = geodata::gadm(country="GUY", level=1, path=tempdir()) |>
  sf::st_as_sf() |> sf::st_cast() |> sf::st_transform(crs_master) |>
  dplyr::rename(State = NAME_1)

aoi    = dplyr::filter(states, State == "Barima-Waini") 

# visualize
tmap::tmap_mode("view")
tmap::tm_shape(states) + tmap::tm_borders(col = "white", lwd = 0.5) +
  tmap::tm_text("State", col = "white", size = 1, alpha = 0.3, just = "bottom") +
  tmap::tm_shape(country) + tmap::tm_borders(col = "white", lwd = 1) +
  tmap::tm_shape(aoi) + tmap::tm_borders(col = "red", lwd = 2) +
  tmap::tm_text("State", col = "red", size = 2) +
  tmap::tm_basemap("Esri.WorldImagery")
```

## Supervised classification

```{r}
library(ForesToolboxRS)
# Data Preparation
dir.create("testdata")
# downloading the image
download.file("https://github.com/ytarazona/ft_data/raw/main/data/LC08_232066_20190727_SR.zip",
              destfile = "testdata/LC08_232066_20190727_SR.zip")
# unziping the image
unzip("testdata/LC08_232066_20190727_SR.zip", exdir = "testdata")
# downloading the signatures
download.file("https://github.com/ytarazona/ft_data/raw/main/data/signatures.zip",
              destfile = "testdata/signatures.zip")
# unziping the signatures
unzip("testdata/signatures.zip", exdir = "testdata")


# Read raster
image <- stack("testdata/LC08_232066_20190727_SR.tif")

# Read signatures
sig <- read_sf("testdata/signatures.shp")

# Classification with Random Forest
classRF <- mla(img = image, model = "randomForest", endm = sig, training_split = 80)

# Results
print(classRF)
```

```{r}
# Classification
colmap <- c("#0000FF","#228B22","#FF1493", "#00FF00")
plot(classRF$Classification, main = "RandomForest Classification", col = colmap, axes = TRUE)#
```

## Monte Carlo Cross-Validation

```{r}
cal_ml <- calmla(img = image, endm = sig,
                 model = c("svm", "randomForest", "naiveBayes", "knn"),
                 training_split = 70, approach = "MCCV", iter = 10)
```

```{r}
# Calibration result
plot(
  cal_ml$svm_mccv,
  main = "Monte Carlo Cross-Validation calibration",
  col = "darkmagenta",
  type = "b",
  ylim = c(0, 0.4),
  ylab = "Error between 0 and 1",
  xlab = "Number of iterations"
)
lines(cal_ml$randomForest_mccv, col = "red", type = "b")
lines(cal_ml$naiveBayes_mccv, col = "green", type = "b")
lines(cal_ml$knn_mccv, col = "blue", type = "b")
legend(
  "topleft",
  c(
    "Support Vector Machine",
    "Random Forest",
    "Naive Bayes",
    "K-nearest Neighbors"
  ),
  col = c("darkmagenta", "red", "green", "blue"),
  lty = 1,
  cex = 0.7
)
```

Monte Carlo of activity data tainty of LULC classification maps and activity data.

This section showcases application of Monte Carlo simulations to uncer-

Import data

remotes::install_github("ytarazona/ForesToolboxRS")

library(ForesToolboxRS)

dir.create("./data/testdata")

download.file("https://github.com/ytarazona/ft_data/raw/main/data/LC08_232066_20190727_SR.zip", destfile

unzip("testdata/LC08_232066_20190727_SR.zip", exdir = "testdata")

download.file("https://github.com/ytarazona/ft_data/raw/main/data/signatures.zip", destfile = "testdata/

unzip("testdata/signatures.zip", exdir = "testdata")

image \<- stack("./data/testdata/LC08_232066_20190727_SR.tif")

sig \<- read_sf("./data/testdata/signatures.shp")

RandomForest classifier

classRF \<- mla(img = image, model = "randomForest", endm = sig, training_split = 80)

print(classRF)

Classify land cover

\# Classification

colmap \<- c("#0000FF","#228B22","#FF1493", "#00FF00")

plot(classRF\$Classification, main = "RandomForest Classification", col = colmap, axes = TRUE)

8Figure

2: LULC map classified with randomForest classifier kernel

Monte Carlo of classification uncertainty

plot(

cal_ml\$svm_mccv,

main = "Monte Carlo Cross-Validation calibration",

col = "darkmagenta",

type = "b",

ylim = c(0, 0.4),

ylab = "Error between 0 and 1",

xlab = "Number of iterations"

)

lines(cal_ml\$randomForest_mccv, col = "red", type = "b")

lines(cal_ml\$naiveBayes_mccv, col = "green", type = "b")

lines(cal_ml\$knn_mccv, col = "blue", type = "b")

legend(

"topleft",

c(

"Support Vector Machine",

"Random Forest",

"Naive Bayes",

"K-nearest Neighbors"

),

lty = 1,

col = c("darkmagenta", "red", "green", "blue"),

9cex = 0.7

)

Figure

3: Monte Carlo ensemble of uncertainty estimates of common classifiers.
